---
title: "Biomarker-Stratified Treatment Effect in Real-World Colon Cancer Data"
subtitle: "From cohort to subgroup forest plot — a reproducible RWE Analysis Report"
author: "Natália S Kugel, PhD"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    theme: default
    embed-resources: true
bibliography: references.bib
---

```{r setup}
#| include: false
library(rwbiomarker)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(gtsummary)
library(broom)

set.seed(2026)
theme_set(theme_minimal(base_size = 12))
```

# Background and objective

Precision medicine asks a deceptively simple question: **does this
treatment work better in some patients than in others?**  In a regulated
RWE submission, answering this question requires a transparent and
reproducible chain of evidence: a clearly specified cohort, a
pre-defined biomarker, pre-specified subgroups, and statistical methods
that handle the time-to-event nature of the data correctly.

In this report we use the publicly available **NCCTG colon adjuvant
chemotherapy cohort** (`survival::colon`) [@moertel1990levamisole] as a
stand-in for an EHR-derived oncology dataset. The trial compared
observation versus levamisole + 5-FU in resected colon cancer. We
augment the cohort with a *simulated* tumor biomarker (think MSI status
or a gene-expression signature) and ask:

> **Does the treatment effect on overall survival differ between
> biomarker-positive and biomarker-negative patients?**

::: {.callout-note}
## Content

By the end of this report you should be able to:

1.  Load and clean a real-world cohort.
2.  Stratify a cohort by a biomarker.
3.  Estimate Kaplan–Meier survival curves and Cox HRs by subgroup.
4.  Test a treatment-by-biomarker interaction.
5.  Visualize subgroup HRs in a forest plot.
6.  Interpret the results in a regulatory context.
7.  Discuss the limitations and what a regulator would want to see next.
:::

# Cohort

```{r cohort}
cohort <- load_colon() |> add_biomarker(prevalence = 0.35, seed = 7)
glimpse(cohort)
```

```{r tbl1}
cohort |>
  select(treatment, biomarker, age, sex, nodes, differ, extent) |>
  tbl_summary(
    by = treatment,
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing = "ifany"
  ) |>
  add_overall() |>
  modify_caption("**Table 1. Baseline characteristics by treatment arm.**")
```

# Overall Kaplan–Meier

```{r km-overall}
#| label: fig-km-overall
#| fig-cap: "**Overall survival by treatment arm.** Kaplan–Meier estimates of overall survival in the full cohort (n = 619; 291 deaths) comparing adjuvant Lev+5FU (orange) versus observation (blue). Shaded bands denote pointwise 95% confidence intervals; the risk table below the curves reports the number of patients still at risk at each timepoint. The Lev+5FU curve lies consistently above the observation curve from approximately one year onward, indicating a sustained survival advantage. This visual separation is consistent with the unadjusted hazard ratio of 0.69 (95% CI 0.55–0.87; p = 0.002) reported in Section *Subgroup hazard ratios*. Time is expressed in days since enrollment."
#| fig-width: 8
#| fig-height: 6
fit <- survfit(Surv(time, status) ~ treatment, data = cohort)
ggsurvplot(fit, data = cohort, conf.int = TRUE, risk.table = TRUE,
           palette = c("#4C78A8", "#F58518"),
           legend.labs = c("Obs", "Lev+5FU"),
           xlab = "Days since enrollment", ylab = "Overall survival")
```

# Biomarker-stratified Kaplan–Meier

```{r km-strat}
#| label: fig-km-strat
#| fig-cap: "**Overall survival by treatment arm, stratified by biomarker status.** Kaplan–Meier curves for Lev+5FU (orange) versus observation (blue) in biomarker-positive (left; n = 217, 102 events) and biomarker-negative (right; n = 402, 189 events) strata. Shaded bands denote 95% pointwise confidence intervals. Curve separation is visually similar across strata, consistent with subgroup HRs of 0.70 (95% CI 0.48–1.04) and 0.68 (95% CI 0.51–0.91) and a non-significant pre-specified treatment-by-biomarker interaction test (LRT p = 0.89). The wider confidence bands in the biomarker-positive panel reflect the smaller number of events rather than a diminished treatment effect."
#| fig-width: 10
#| fig-height: 5
fit_pos <- survfit(Surv(time, status) ~ treatment,
                   data = filter(cohort, biomarker == "Positive"))
fit_neg <- survfit(Surv(time, status) ~ treatment,
                   data = filter(cohort, biomarker == "Negative"))

p_pos <- ggsurvplot(fit_pos, data = filter(cohort, biomarker == "Positive"),
                    conf.int = TRUE, palette = c("#4C78A8", "#F58518"),
                    title = "Biomarker positive",
                    legend.labs = c("Obs", "Lev+5FU"))$plot
p_neg <- ggsurvplot(fit_neg, data = filter(cohort, biomarker == "Negative"),
                    conf.int = TRUE, palette = c("#4C78A8", "#F58518"),
                    title = "Biomarker negative",
                    legend.labs = c("Obs", "Lev+5FU"))$plot
gridExtra::grid.arrange(p_pos, p_neg, ncol = 2)
```

# Subgroup hazard ratios

```{r hr-subgroup}
sg <- fit_subgroup_hr(cohort) |>
  filter(term == "treatmentLev+5FU") |>
  transmute(subgroup, n, events,
            HR = round(estimate, 2),
            `95% CI` = sprintf("%.2f – %.2f", conf.low, conf.high),
            `p-value` = signif(p.value, 3))
knitr::kable(sg,
             caption = "Treatment effect (Lev+5FU vs Obs) on OS by biomarker stratum.")
```


**Interpretation:** 
Adjuvant Lev+5FU was associated with a statistically
significant reduction in the hazard of death in the overall cohort
(HR 0.69, 95% CI 0.55–0.87; p = 0.0017), corresponding to an estimated
31% relative risk reduction versus observation. Subgroup-specific
estimates were essentially identical in biomarker-positive
(HR 0.70, 95% CI 0.48–1.04) and biomarker-negative
(HR 0.68, 95% CI 0.51–0.91) patients, with fully overlapping confidence
intervals. The wider interval and larger p-value in the
biomarker-positive subgroup reflect the smaller number of events
(n = 102 vs. 189) rather than a smaller treatment effect; the difference
in subgroup p-values is therefore **not** evidence of effect
modification. Formal assessment of a predictive biomarker effect is
deferred to the pre-specified treatment-by-biomarker interaction test
(Section *Treatment × biomarker interaction*), in line with regulatory
guidance that subgroup p-value comparisons should not be used to infer
heterogeneity of treatment effect.


## Forest plot

```{r forest}
#| label: fig-forest
#| fig-cap: "**Forest plot of treatment effect on overall survival by biomarker subgroup.** Hazard ratios (HR) for Lev+5FU vs. observation are shown on a logarithmic scale; horizontal lines denote 95% confidence intervals. Point estimates were near-identical across biomarker-positive (HR 0.70) and biomarker-negative (HR 0.68) strata, with overlapping confidence intervals, consistent with the non-significant pre-specified interaction test (LRT p = 0.89). The wider interval in the biomarker-positive subgroup reflects fewer events (102 vs. 189) rather than a smaller treatment effect."
#| fig-width: 10
#| fig-height: 5
plot_data <- fit_subgroup_hr(cohort) |>
  filter(term == "treatmentLev+5FU") |>
  mutate(label = sprintf("%s (n=%d, events=%d)", subgroup, n, events))

ggplot(plot_data,
       aes(y = reorder(label, estimate),
           x = estimate, xmin = conf.low, xmax = conf.high)) +
  geom_pointrange(color = "#1f77b4", size = 0.6) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(x = "Hazard ratio (log scale)", y = NULL,
       title = "Forest plot: treatment effect by biomarker subgroup",
       subtitle = "HR < 1 favors Lev+5FU")
```

# Treatment × biomarker interaction

```{r interaction}
out <- fit_interaction_cox(cohort)
out$coefs |>
  transmute(term, HR = round(estimate, 2),
            `95% CI` = sprintf("%.2f – %.2f", conf.low, conf.high),
            `p-value` = signif(p.value, 3)) |>
  knitr::kable(caption = "Adjusted Cox model with treatment x biomarker interaction.")
```

The likelihood-ratio test p-value for the interaction is
**`r signif(out$lrt_p, 3)`**.


**Interpretation.** 
The pre-specified treatment-by-biomarker interaction was not
significant (HR 0.97, 95% CI 0.59–1.59; LRT p = 0.89), providing
no evidence of a predictive biomarker effect; the overall survival
benefit of Lev+5FU was consistent across biomarker strata.


# Limitations and discussion

- The biomarker here is **simulated**; in a real submission, biomarker
  assay validation and pre-specification are essential.
- The colon dataset is from a randomized trial — for a true RWE study,
  one would also handle confounding (e.g., propensity score adjustment).
- Multiplicity: subgroup HRs and the interaction test should be
  interpreted with appropriate adjustment.
- The proportional-hazards assumption should be checked
  (`cox.zph`) — left as an exercise.

# Reproducibility

```{r repro}
sessionInfo()
```

# References {.unnumbered}

::: {#refs}
:::
